---
title: "Week 14 Classwork/Homework Assignment"
author: "John Vinson"
date: "11/13/2017"
output: pdf_document
---

*For the assignment, you will submit the compiled pdf document as both a pdf. Make sure that the compiled documents will display all the required code to get your results.*

\begin{center}
{\bf The assignment is due Tuesday November 28 at 3:30 PM.}
\end{center}

\vspace{24pt}

1. Compare and contrast stochastic and deterministic models. Make sure you discuss what they do and assume, when they are appropriate to use and 

\vspace{12pt}

2. Outline Gillespie’s direct method algorithm.

\vspace{12pt}

3. Assuming a deterministic model plot the trajectory of the density-dependent population model. What happens to the population over time. 

\begin{equation}
\frac{dN}{dt} = (b_0 - b_1N)N - (d_0 + d1N)N 
\end{equation}

Using the parameters and initial conditions

Table 1. Parameter values and initial condition for questions 3 and 5. 
\begin{center}
```{r, echo=FALSE, results="asis", fig.align='center'}
library("knitr")

parms = c("b0", "b1", "d0", "d1", "N0")
vals = c(1.0, 0.001, 0.5, 0.004, 1)

le.df = data.frame(cbind(parms, vals))

kable(le.df, col.names = c("Parameter", "Value"), format="latex", align="c")
```
\end{center}

```{r, echo=FALSE, eval=FALSE}
library(deSolve)

pop = function(t, y, parms){
  with(as.list(c(y, parms)), {
    
    dN = (b0-b1*N)*N - (d0+d1*N)*N 
    
    return(list(c(dN)))
  })
}

time.vec = seq(0, 100, 0.1)

b0 = 1.0
b1 = 0.001

d0 = 0.5
d1 = 0.004

params = c(b0, b1, d0, d1)

N0 = 1
y0 = c(N = N0)

out = data.frame(ode(y0, time.vec, pop, params))

plot(N~time, data=out)
```

\vspace{12pt}
4. Convert the deterministic model into a stochastic model using Gillespie’s direct method algorithm. 

```{r, echo=FALSE, eval=FALSE}
pop.onestep = function(x, params){
  N = x[2]
  
  with(as.list(params), 
       {
         rates = c((b0-b1*N)*N, (d0+d1*N)*N)
         
         changes = matrix(c(1, -1), ncol=1, byrow=TRUE)
         
         tau = rexp(1, rate=sum(rates))
         x[1] = x[1] + tau
         
         U = runif(1)
         
         m = min(which(cumsum(rates)>=U*sum(rates)))
         
         x[2] = x[2] + changes[m,]
         
         return(out = x)
       }
  )
}


pop.model = function(x0, params, nstep){
  output = array(dim=c(nstep+1, 2))
  colnames(output) = c("time", "N")
  x = c(0, x0)
  output[1,] = x

  for(i in 1:nstep){
    output[i+1,] = x = pop.onestep(x, params)
  }
  return(output)
}

b0 = 1.0
b1 = 0.001

d0 = 0.5
d1 = 0.004

params = c(b0, b1, d0, d1)

N0 = 1
y0 = c(N = N0)

out.stoc = pop.model(y0, params, 200)

```

\vspace{12pt}

5. Plot the trajectory of 50 simulations using the parameters found in Table 1. For each simulation allow the simulation to run for 500 steps.
\vspace{12pt}

6. Modify the algorithm to no longer require a number of time steps but rather will end the simulation when the population goes extinct. 

\vspace{12pt}

7. Determine the probability that a population goes extinct within 50 time steps using the parameters found in table 2? At any time?

Table 2. Parameter values and initial conditions for question 7. 
\begin{center}
```{r, echo=FALSE, results="asis", fig.align='center'}
library("knitr")

parms = c("b0", "b1", "d0", "d1", "N0")
vals = c(1.0, 0.001, 0.5, 0.004, 1)

le.df = data.frame(cbind(parms, vals))

kable(le.df, col.names = c("Parameter", "Value"), format="latex", align="c")
```
\end{center}

```{r, echo=FALSE, eval=FALSE}
nsim = 1000

pop.model.new = function(x0, params){
  output = NULL
  x = c(0, x0)
  output = matrix(x, nrow=1)
  colnames(output) = c("time", "N")

  while(x[2]>0){
    x = pop.onestep(x, params)
    output = rbind(output, x)
  }
  return(output)
}

b0 = 1.0
b1 = 0.001

d0 = 1.0
d1 = 0.004

params = c(b0, b1, d0, d1)

out.new = pop.model.new(y0, params)

```

\vspace{12pt}

8. Using the zombie outbreak model from the previous case studies, write a stochastic model.  Develop and address 2 problems or questions using this model. 

\begin{align}
\frac{dS}{dt} =& bS - mS - aSZ\\
\frac{dU}{dt} =& aSZ - zU\\
\frac{dZ}{dt} =& zU - kSZ\\
\frac{dD}{dt} =& kSZ + mS
\end{align}
