---
title: "Week 3 Class Problems"
author: "John Vinson"
date: "8/29/2017"
output: pdf_document
---

1. For many species of fish, the weight W (in g) is a function of the length L (in mm) that can be expressed by

\begin{equation}
W = W(L) = kL^3
\end{equation}

where k is a constant. For a particular species of fish $k=0.02$ and the length of this species is a function of the number of years that the fish has been alive, t, and is given by

\begin{equation}
L = L(t) = 50-\frac{(t-20)^2}{10}
\end{equation}
with no fish living longer than 20 years. 

- Determine how the weight of the fish changes with their age (time that they've been alive).
- What are all the possible values of k? 
- Plot how the weight of the fish changes through their lifetime. 
- Mathematically find the age of the fish for which the weight will exceed 450 g.
- Numerically find the age of the fish for which the weight will exceed 450 g. 

```{r, echo=F, eval=FALSE}
W = function(t){
  L = 50 - ((t-20)^2)/10
  return(0.02*L^3)
}

t.vec = seq(0, 20)

plot(t.vec, W(t.vec))
```

\vspace{12pt}

2. For fiddler crabs, data gathered by Thompson show that the relationship between the weight C of the claw and the weight of the body W is given by

\begin{equation}
C = 0.11W^{1.54}
\end{equation}
where W and C are in grams.  

- What values can C and W take? 
- Plot C for $W \in (0, 20]$.  
- Write the function where the weight of the crab body is a function of claw weight.
- For what values of the claw weight does the crab body weight exceed 6 g?

```{r, echo=FALSE, eval=FALSE}
fc = function(W){
  return(0.11*W^1.54)
}

W.vec = seq(0,20, 0.1)

plot(W.vec, fc(W.vec))
```

\begin{figure}
\centerline{\includegraphics[width=0.5\textwidth]{/opt/home/vinsonje/ECOL3900_Vinson_F17/fiddler_crab.png}}
\caption{Fiddler crab, \textit{Ucides cordatus}, from Margrave (1648)}
\end{figure}

\vspace{24pt}

3. The following table give the number of AIDS-related deaths in the New York Prison System.

```{r, echo=FALSE, results="asis", fig.align='center'}
library("knitr")

years = seq(1981, 1989, 1)
years.2 = seq(1990, 1998, 1)
deaths = c(2,4,18,57,99,124,151,158, 132)
deaths.2 = c(175,229,208,223,246,256,181,60,39)

aids.df = data.frame(years, deaths, years.2, deaths.2)

kable(aids.df, col.names = c("year", "deaths", "year", "deaths"), format="latex", align="c")
```

- Create a scatter plot for the data in the table. Let the x-axis represent the number of years since 1980.
- Determine an appropriate function that can be used to model the data. Why did you choose this function? Estimate the parameter values.
- Using this model(function), determine what year the deaths was a maximum.

```{r, echo=FALSE, eval=FALSE}
years = seq(1981,1998)-1980
deaths = c(2,4,18,57,99,124,151,158, 132,175,229,208,223,246,256,181,60,39)

aids.data = data.frame(years, deaths)
cub.fun = function(parms, year){return(
  parms[1]*year^3 + parms[2]*year^2 + parms[3]*year + parms[4]
)}

sse = function(parms, data){
  out = sum((cub.fun(parms, data[,1])-data[,2])^2)
  return(out)
}

out = optim(c(0,0,0,0), fn=sse, data=aids.data, method="Nelder-Mead")

plot(deaths~years)
lines(years, cub.fun(out$par, years))

book.ans = c(-0.278,5.543,-7.898,9.101)
lines(years, cub.fun(book.ans, years), col="red")

sse(book.ans, aids.data)
sse(out$par, aids.data)
```
